Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-08-14 grove-paper.tex
More work on "Add" and "Change" sections and some TikZ improvements

\usetikzlibrary{
  graphs,       % Graph *notation*
  graphdrawing, % Graph *layout*
}

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:A simple tree with a hole-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> [] {
   22[> edge label={left}, > edge label'={2}, as={$\texttt{var(x)}_3$}],
   {}
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:A simple tree with a hole-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}

@@

Now that we have discussed adding holes

\subsubsection{Add}

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:Addition-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   x [> edge label={left}, > edge label'={2}, as={$\texttt{var(x)}_2$}],
   y [> edge label={right}, > edge label'={3}, as={$\texttt{var(y)}_3$}]
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:Addition-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Addition}
\label{fig:Addition}
\end{figure}

We model edits to the code as sequences of additions and deletions of edges to this graph.
For example, consider the graph in Figure~\ref{fig:Delete and Add}, which
represents the expression \texttt{1 + 2}.
Suppose we want to change the \texttt{2} into a \texttt{3} so the expression
becomes \texttt{1 + 3}.

When users insert code into a hole, our graph model simply adds a corresponding a vertex and edge.
For example, suppose the user adds \texttt{y} as the right child of vertex~1 in Figure~\ref{fig:A simple tree with a hole (Alternate figure)}.
This corresponds to creating a vertex for the variable reference \texttt{y} and adding an edge
to that vertex from vertex~1.
Figure~\ref{fig:Addition} shows the result, where the new vertex has identifier~3.

Note that adding an edge implicitly creates.

\subsubsection{Change}

In order to change an existing vertex, we delete the existing edge to that vertex and add the new value.
For example, in

Note that while we
Note that the vertex continues to exist (this is essential for further manipulation)
Once created a vertex is never deleted.

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   x [> edge label={left}, as={$\texttt{var(x)}_2$}],
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_3$}]
};
\end{tikzpicture}
\caption{TODO}
\label{fig:Change-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   x [> edge label={left}, as={$\texttt{var(x)}_2$}],
   z [> edge label={right}, as={$\texttt{var(z)}_4$}]
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_3$}]
};
\end{tikzpicture}
\caption{TODO}
\label{fig:Change-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Change}
\label{fig:Change}
\end{figure}

\begin{figure}
\centering
\begin{tikzpicture}
\path (0,0)    node {\includegraphics[scale=0.75]{graphviz/delete-and-add/1.dot}};
\path (0,-4cm) node {\includegraphics[scale=0.75]{graphviz/delete-and-add/2.dot}};
\path (0,-8cm) node {\includegraphics[scale=0.75]{graphviz/delete-and-add/3.dot}};
\end{tikzpicture}
\caption{Delete and Add}
\label{fig:Delete and Add}
\end{figure}
NEEDS UI EXAMPLE

\footnote{For the purposes of this paper, we represent identifiers and number literals
indivisibly.  See Section~TODO:REF, for how we would allow editing individual characters.}

We model this by first deleting edge~7.
This results in the middle graph in Figure~\ref{fig:Delete and Add}, which
represents the expression \texttt{1 + \hole}.
Note that vertex~9 still exists and can be manipulated (which we will exploit in the next subsection).
Vertex~9 is just not reachable from the root.

After that deletion, we then add a new vertex to represent~\texttt{3} and an
edge to it.  This results in the bottom graph in Figure~\ref{fig:Delete and Add}.

@@

TODO: move subsubsection elsewhere?

>>>>

\usetikzlibrary{
  graphs,       % Graph *notation*
  graphdrawing, % Graph *layout*
  quotes,       % Quote syntax (e.g., "foo")
}

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:A simple tree with a hole-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   {}
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:A simple tree with a hole-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}

@@

\subsubsection{Additions}

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:Addition-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   "$\texttt{var(y)}_5$" [> "right", > "4"']
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:Addition-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Addition}
\label{fig:Addition}
\end{figure}

We model edits to the code as sequences of additions and deletions of edges to this graph.
For example, suppose the user adds \texttt{y} as the right child of vertex~1 in Figure~\ref{fig:A simple tree with a hole (Alternate figure)}
so that the expression becomes \texttt{x + y}.
This corresponds to creating a vertex for the variable reference \texttt{y} and adding an edge
to that vertex from vertex~1.
Figure~\ref{fig:Addition} shows the result, where the new vertex has identifier~3.

TODO: Note that adding an edge implicitly creates.

For the purposes of this paper, we represent identifiers and number literals
indivisibly.  See Section~TODO:REF, for how we would allow editing individual characters.

\subsubsection{Changes}

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   {}
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_3$}]
};
\end{tikzpicture}
\caption{TODO}
\label{fig:Change-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   "$\texttt{var(z)}_7$" [> "right", > "6"']
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 "$\texttt{var(y)}_5$" []
};
\end{tikzpicture}
\caption{TODO}
\label{fig:Change-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Change}
\label{fig:Change}
\end{figure}

% \begin{figure}
% \centering
% \begin{tikzpicture}
% \path (0,0)    node {\includegraphics[scale=0.75]{graphviz/delete-and-add/1.dot}};
% \path (0,-4cm) node {\includegraphics[scale=0.75]{graphviz/delete-and-add/2.dot}};
% \path (0,-8cm) node {\includegraphics[scale=0.75]{graphviz/delete-and-add/3.dot}};
% \end{tikzpicture}
% \caption{Delete and Add}
% \label{fig:Delete and Add}
% \end{figure}

Next suppose we want to replace \texttt{y} with \texttt{z} so that the code becomes \texttt{x + z}.
This is accomplished by first deleting edge~4.
This results in Figure~\ref{fig:Change-a}, which represents the expression \texttt{x + \hole{}}.
After that deletion, we add a new vertex to represent~\texttt{z} and an
edge from vertex~1 to that vertex.
This results in the bottom graph in Figure~\ref{fig:Change-b},
which represents the expression \texttt{x + z}.

Note that vertex~3 continues to exist, and if it had any children, those children could remain connected to it.
In our system, once a vertex is created it is never deleted.
This allows further manipulations of those vertexes as shown later in this section.

@@

@@
